---
title: "Model for ART utilisation over time"
subtitle: "Complete case analysis"
author: "Oisin Fitzgerald"
output:
  html_document:
    toc: true
---

```{r,echo=FALSE}
knitr::opts_chunk$set(echo=FALSE,message=FALSE,warning=FALSE)
```

# Introduction

This document describes the longitudinal modeling of assisted reproductive 
technology utilisation (ART) as a function of the United Nations produced 
Human Development Index (HDI) and Gender Inequality Index (GII). Greater details on the 
rationale for the study can be found in the main manuscript. 

Summary of methods and findings:

* Linear mixed effect model for log ART over time with covariates GII and HDI  
* GII only available for year XXX so data restricted to these years.
* GII related to log ART, increased GII related to increased ART utilisation. Effect marginally significant.

```{r}
### packages, functions
# external
library(lme4)
library(MuMIn)
library(ggplot2)
library(data.table)
library(broom.mixed)
# internal
source("../src/tools/visualise.R")
source("../src/tools/summarise.R")
source("../src/tools/modelling.R")
source("../src/lmm_ridge.R")

if(!dir.exists("../models")) {
  dir.create("../models")
}
plot_resid_against_x <- function(x,augment_model) {
  x <- substitute(x)
  ggplot(augment_model) +
    geom_point(aes_(y=quote(.resid),x=x)) +
    geom_smooth(aes_(y=quote(.resid),x=x)) +
    geom_label(data=augment_model[abs(.resid)>1],
               aes_(y=quote(.resid + 0.2),x=x,label=quote(country))) +
    theme_bw(base_size = 14)
}


qqplot_ranef <- function(model) {
  ranef_model <- data.table(ranef(model)$country,keep.rownames = TRUE)
  names(ranef_model) <- c("country","intercept")
  ggplot(ranef_model) +
    geom_qq(aes(sample=intercept)) +
    geom_abline(slope = 1,intercept=0,linetype=2,col="red") +
    theme_bw(base_size = 14)
}

plot_cooks_dist <- function(augment_model) {
  ggplot(augment_model,aes(x=.fitted,y=.cooksd)) +
  geom_point() +
  geom_label(data=augment_model[abs(.cooksd)>0.5],
             aes(y=(.cooksd + 0.2),x=(.fitted),label=(country)),alpha=0.5) +
  theme_bw(base_size=16)
}

marginal_r2 <- function(model,df,vars,y) {
  X <- as.matrix(df[,vars,with=FALSE])
  yhat <- X %*% model$beta
  msk <- !is.na(yhat) & !is.na(y)
  var(yhat[msk]) / var(y[msk])
}


### import data
art_dat <- readRDS("../data/art_covar_2002_2014.rds")
setDT(art_dat)
```

# Data

There is more detailed description of the data used in the following files:  

* ...

Several countries are removed from the analysis due to poor data quality (or missing data in the case of Taiwan)

```{r}
# country exclusions
countries_remove <- c("SaudiArabia", "UnitedArabEmirates", "Jordan", "Lebanon", "Bahrain",
                       "SyrianArabRepublic", "Morocco",
                       "Tunisia", "Libya", "Palestine,Stateof", "Egypt",
                       "Taiwan")
art_dat[,exclude := fifelse(country %in% countries_remove,TRUE,FALSE,FALSE)]
art_dat[,country := factor(country)]

# middle east, north africa variable
me_na <- c("SaudiArabia", "UnitedArabEmirates", "Jordan", "Lebanon", "Bahrain",
                       "SyrianArabRepublic", "Morocco",
                       "Tunisia", "Libya", "Palestine,Stateof", "Egypt")
art_dat[,me_na := fifelse(country %in% me_na,1,0,0)]
```

```{r}
write.csv(art_dat[exclude == FALSE & !is.na(art_utilisation),.N,by=.(country,region)],
          "../results/countries_raw.csv")
```

# Index models

We create two linear mixed effect models for log ART utilisation. Both include 
random intercept terms for country and time, the HDI and GII as variables. The second 
model differs in decomposing the variable effects into within country and between country
effects. The rationale for this was to assess whether there was evidence that increases
in GII within a country over the timeframe of our dataset were associated with increased 
ART utilisation. 

We note the following:  

* The GII is only available for the years: XXX, 
* Several countries are missing outcome data for some of the relevant year.
* Time is years, centered so that 0 is XXXX.
* The GII and HDI are between 0 and 1.

```{r}
# prepare data - scaling
scale <- function(x) (x - mean(x,na.rm=TRUE))/sd(x,na.rm = TRUE)
art_dat[,time := scale(time)]
art_dat[,gii_un := scale(gii_un)]
art_dat[,hdi_un := scale(hdi_un)]
art_dat[,ln_art_utilisation1 := ln_art_utilisation - mean(ln_art_utilisation,na.rm=TRUE)]
```

## Standard formulation

Model: log ART utilisation = time in years + GII + HDI + random country intercept

```{r}
# model
mm01 <- lmer(ln_art_utilisation ~ 1+time+gii_un+hdi_un+(1|country), 
             data = art_dat[exclude==FALSE])
saveRDS(mm01,file="../models/index_standard_complete.rds")
cat("R2: ",MuMIn::r.squaredGLMM(mm01),"\n")

# model details
augment_mm01 <- augment(mm01)
setDT(augment_mm01)

# model fit
knitr::kable(tidy(mm01),digits = 3,caption = "Model parameter estimates")
knitr::kable(glance(mm01),digits = 3,caption = "Model fit statistics")
knitr::kable(confint(mm01),digits=3,caption = "Model 95% intervals")

fwrite(x = cbind(tidy(mm01)[1:4,],confint(mm01)[3:6,]),file = "../results/tables/index_standard_complete.csv")

```

On the exp scale:

```{r}
knitr::kable(tidy(mm01),digits = 3,caption = "Model parameter estimates")
knitr::kable(confint(mm01),digits=3,caption = "Model 95% intervals")
```

Plot on the exp scale:

```{r}
mm01f <- fixef(mm01)["gii_un"]
mm01vc <- vcov(mm01)["gii_un","gii_un"]
change_gii_un <- seq(0,-1,by=-0.1)
df <- data.frame(change_gii_un,
                 change_art=exp(change_gii_un*mm01f),
                 change_art_lower=exp(change_gii_un*(mm01f-2*mm01vc)),
                 change_art_upper=exp(change_gii_un*(mm01f+2*mm01vc)))
ggplot(df,aes(x=change_gii_un,y=change_art)) +
  geom_line() +
  scale_y_continuous(breaks=seq(1,2,by=0.25),labels=seq(0,100,by=25)) +
  geom_ribbon(aes(ymin=change_art_lower,ymax=change_art_upper),alpha=0.1)+
  labs(x = "Change in Gender Inequality Index\n(SD units)",y="Change in ART utilisation (%)") +
  theme_bw(base_size=14)
ggsave("../results/graphs/change_gii_adjusted.png")
```

```{r}
knitr::kable(df)
```

The quantile-quantile plot for the model appears reasonable.

```{r, fig.height=4, fig.width=8}
qqplot_ranef(mm01)
```

There is evidence of outliers or poor model fit for a couple of south/central American countries.

```{r, fig.height=4, fig.width=8}
plot_resid_against_x(time,augment_mm01)
plot_resid_against_x(`.fitted`,augment_mm01)
plot_resid_against_x(hdi_un,augment_mm01)
plot_resid_against_x(gii_un,augment_mm01)
```

Influence

```{r}
plot_cooks_dist(augment_mm01)
```

## Standard formulation - each index individually (unadjusted)

Model: log ART utilisation = time in years + GII + random country intercept

```{r}
# model
model <- lmer(ln_art_utilisation ~ 1+time+gii_un+hdi_un+(1|country), 
             data = art_dat[exclude==FALSE])
saveRDS(model,file="../models/index_gii_complete.rds")
cat("R2: ",MuMIn::r.squaredGLMM(model),"\n")

# model details
augment_model <- augment(model)
setDT(augment_model)

# model fit
knitr::kable(tidy(model),digits = 3,caption = "Model parameter estimates")
knitr::kable(glance(model),digits = 3,caption = "Model fit statistics")
knitr::kable(confint(model),digits=3,caption = "Model 95% intervals")
```

Model: log ART utilisation = time in years + HDI + random country intercept

```{r}
# model
model <- lmer(ln_art_utilisation ~ 1+time+hdi_un+(1|country), 
             data = art_dat[exclude==FALSE])
saveRDS(model,file="../models/index_hdi_complete.rds")
cat("R2: ",MuMIn::r.squaredGLMM(model),"\n")

# model details
augment_model <- augment(model)
setDT(augment_model)

# model fit
knitr::kable(tidy(model),digits = 3,caption = "Model parameter estimates")
knitr::kable(glance(model),digits = 3,caption = "Model fit statistics")
knitr::kable(confint(model),digits=3,caption = "Model 95% intervals")
```

## Decomposition formulation

This model decompose effects effects into within country and between country effects. 
effects. The resulting between and within coefficients can be compared to assess whether 
it is reasonable to assume that the between and within country effects are similar, which 
can be used to reason about causation of the effect etc. 
We originally decomposed time as it is measured
unevenly between countries (due to missing outcome data) and this is recommended in the 
literature, however it results in non-identifiability.

In general the previous model is a better fit, with better model fit statistics and 
a hypothesis test that the between and within effects of GII are equivalent being 
non-significant. 

```{r}
# between within decomposition
art_dat[,time_m := mean(time,na.rm=TRUE),by=country]
art_dat[,time_d := time - time_m]
art_dat[,gii_m := mean(gii_un,na.rm=TRUE),by=country]
art_dat[,gii_d := gii_un - gii_m]
art_dat[,hdi_m := mean(hdi_un,na.rm=TRUE),by=country]
art_dat[,hdi_d := hdi_un - hdi_m]

# model
mm02 <- lmer(ln_art_utilisation ~ 1+time+gii_d+gii_m+hdi_d+hdi_m+(1|country),data = art_dat[exclude==FALSE])
saveRDS(mm02,file="../models/index_decomposition_complete.rds")
cat("R2: ",MuMIn::r.squaredGLMM(mm02),"\n")

# model details
augment_mm02 <- augment(mm02)
setDT(augment_mm02)

# model fit
knitr::kable(tidy(mm02),digits = 3,caption = "Model parameter estimates")
knitr::kable(glance(mm02),digits = 3,caption = "Model fit statistics")
knitr::kable(confint(mm02),digits=3,caption = "Model 95% intervals")


```


As before, the quantile-quantile plot for the model appears reasonable.

```{r, fig.height=4, fig.width=8}
qqplot_ranef(mm02)
```

As before, there is evidence of outliers or poor model fit for a couple of south/central American countries.

```{r, fig.height=4, fig.width=8}
plot_resid_against_x(`.fitted`,augment_mm02)
plot_resid_against_x(hdi_m,augment_mm02)
plot_resid_against_x(gii_m,augment_mm02)
plot_resid_against_x(hdi_d,augment_mm02)
plot_resid_against_x(gii_d,augment_mm02)
```

Influence

```{r}
plot_cooks_dist(augment_mm02)
```

GII only

```{r}
# model
model <- lmer(ln_art_utilisation ~ 1+time+gii_d+gii_m+(1|country),data = art_dat[exclude==FALSE])
cat("R2: ",MuMIn::r.squaredGLMM(model),"\n")

# model details
augment_model <- augment(model)
setDT(augment_model)

# model fit
knitr::kable(tidy(model),digits = 3,caption = "Model parameter estimates")
knitr::kable(glance(model),digits = 3,caption = "Model fit statistics")
knitr::kable(confint(model),digits=3,caption = "Model 95% intervals")
```

```{r, fig.height=4, fig.width=8}
qqplot_ranef(model)
```

```{r, fig.height=4, fig.width=8}
plot_resid_against_x(`.fitted`,augment_model)
plot_resid_against_x(gii_m,augment_model)
plot_resid_against_x(gii_d,augment_model)
```

Influence

```{r}
plot_cooks_dist(augment_model)
```

**This table appears in paper supplements**

```{r}
tab <- tidy(model)
tab <- cbind(tab, broom::confint_tidy(model))
data.table::fwrite(tab,file = "../results/tables/gii_only_longitudinal.csv")
```

# Component models

HDI
* Life expectancy
* Expected years of schooling
* Mean years of schooling 
* GNI per capita

GII 
* Maternal mortality ratio
* Adolescent birth rate  
* Female/male population with at least secondary education 
* Female/male share of parliamentary seats  
* Female/male labour force participation  

## Correlation

Strong correlations between the components.

```{r, fig.height=4, fig.width=8}
vars <- c("ln_adol_birth_rate_wb", "women_parl_wb", "ln_maternal_mortality_wb",
          "school_gap", "labour_gap_wb", "life_expect_wb", "ln_gni_wb", "mean_sch_un")
cor_gii_hdi <- cor(art_dat[,..vars], use="pair", method="pearson")
rownames(cor_gii_hdi) <- colnames(cor_gii_hdi) <- c("Adol. birth rate", "Women in parliament (%)", 
                                                "Maternal mort. ratio","Gender highschool gap", 
                                                "Gender LFP gap","Life expectancy", 
                                                "log GNI", "Av. years of schooling")
corrplot::corrplot(cor_gii_hdi, method = "square", type = "upper", order = "FPC", tl.srt = 50, 
                   diag = FALSE, tl.col = "black", tl.cex = 1.3)
```

Poorly condition matrix.

```{r}
X <- as.matrix(na.omit(art_dat[,..vars]))
eigen_XtX <- eigen(t(X) %*% X)$values
cat("max/min eigenval of XtX:",max(eigen_XtX)/min(eigen_XtX))
```

The cumulative sum of the eigenvalues of the correlation matrix dividing by their sum.

```{r}
vars <- c("time","ln_adol_birth_rate_wb", "women_parl_wb", "ln_maternal_mortality_wb",
          "school_gap", "labour_gap_wb", "life_expect_wb", "ln_gni_wb", "mean_sch_un")
cor_gii_hdi <- cor(art_dat[,..vars], use="pair", method="pearson")
eigen_cor_gii_hdi <- eigen(cor_gii_hdi)
df_plot <- data.frame(var_exp = cumsum(eigen_cor_gii_hdi$values)/sum(eigen_cor_gii_hdi$values),
                      comp = 1:length(vars))
ggplot(df_plot,aes(y=100*var_exp,x=comp)) +
  geom_point() +
  geom_line() +
  theme_bw(base_size=12) +
  labs(x = "Number of components",y = "Variance explained (%)") +
  scale_x_continuous(breaks=1:length(vars))
```

## Standard formulation - no shrinkage on fixed effects

This standard LMM model is overly influence by multicollinearity, fitted below for reference.

```{r}
# model
mc01 <- lmer(ln_art_utilisation ~ 1+
               time+
               ln_adol_birth_rate_wb+
               ln_maternal_mortality_wb+
               women_parl_wb+
               school_gap+
               labour_gap_wb+
               mean_sch_un+
               life_expect_wb+
               ln_gni_wb+
               (1|country),
             data = art_dat[exclude==FALSE],
             REML=FALSE)
saveRDS(mc01,file="../models/component_standard_l20_complete.rds")
cat("R2: ",MuMIn::r.squaredGLMM(mc01),"\n")

# model details
augment_mc01 <- augment(mc01)
setDT(augment_mc01)

# model fit
knitr::kable(tidy(mc01),digits = 3,caption = "Model parameter estimates")
knitr::kable(confint(mc01),digits=3,caption = "Model 95% intervals")
knitr::kable(glance(mc01),digits = 3,caption = "Model fit statistics")
```

## Standard formulation - L2 shrinkage on fixed effects

In order to deal with the collinearity we apply $L2$ shrinkage to the fixed effects
(as is already done to the random effects). We scale the data so the covariates
are zero mean and standard deviation one. We also mean centre the outcome to remove 
the need for an intercept term. 

```{r}
# prepare data - scaling
scale <- function(x) (x - mean(x,na.rm=TRUE))/sd(x,na.rm = TRUE)
#art_dat[,time1 := scale(time)]
art_dat[,time1 := time]
art_dat[,ln_adol_birth_rate_wb1 := scale(ln_adol_birth_rate_wb)]
art_dat[,women_parl_wb1 := scale(women_parl_wb)]
art_dat[,ln_maternal_mortality_wb1 := scale(ln_maternal_mortality_wb)]
art_dat[,school_gap1 := scale(school_gap)]
art_dat[,labour_gap_wb1 := scale(labour_gap_wb)]
art_dat[,life_expect_wb1 := scale(life_expect_wb)]
art_dat[,ln_gni_wb1 := scale(ln_gni_wb)]
art_dat[,mean_sch_un1 := scale(mean_sch_un)]

Xvars1 <- c("time1","ln_adol_birth_rate_wb1", "women_parl_wb1", "ln_maternal_mortality_wb1",
          "school_gap1", "labour_gap_wb1", "life_expect_wb1", "ln_gni_wb1", "mean_sch_un1")
Xvars1_labels <- c("Year","Adol. birth rate","Women in parliament (%)","Maternal mort. ratio",
                   "Gender highschool gap","Gender LFP gap","Life expectancy",
                   "log GNI","Av. years of schooling")
art_dat[,ln_art_utilisation1 := ln_art_utilisation - mean(ln_art_utilisation,na.rm=TRUE)]
```

Firstly, we refit the unpenalised model to the scaled data.

```{r}
# model
mc01s <- lmer(ln_art_utilisation1 ~ -1+
               time1+
               ln_adol_birth_rate_wb1+
               ln_maternal_mortality_wb1+
               women_parl_wb1+
               school_gap1+
               labour_gap_wb1+
               mean_sch_un1+
               life_expect_wb1+
               ln_gni_wb1+
               (1|country),
             data = art_dat[exclude==FALSE],
             REML=FALSE)
cat("R2: ",MuMIn::r.squaredGLMM(mc01s),"\n")

# model details
augment_mc01s<- augment(mc01s)
setDT(augment_mc01s)

# model fit
knitr::kable(tidy(mc01s),digits = 3,caption = "Model parameter estimates")
knitr::kable(confint(mc01s),digits=3,caption = "Model 95% intervals")
knitr::kable(glance(mc01s),digits = 3,caption = "Model fit statistics")
```

Fitting the model over various size of $\lambda$ in the $L2$ penalty $\lambda\| \beta \|^2$ 
gives the results shown in the graph below. A $\lambda$ corresponding to an effective 
number of parameters of 5 appears a reasonable choice in stabilising the coefficient values. 

```{r}
Lambda <- seq(1,1500,length.out = 20)
effective_df <- numeric(length=length(Lambda))
X <- as.matrix(na.omit(art_dat[exclude==FALSE,..Xvars1]))
for (i in 1:length(effective_df)) {
  effective_df[i] <- sum(diag(X %*% solve(t(X) %*% X + diag(Lambda[i],length(Xvars1))) %*% t(X)))
}

betas <- data.frame(beta=numeric(length = length(Lambda)*length(Xvars1)),
                    var=rep(Xvars1,length(Lambda)),
                    lambda=rep(Lambda,each=length(Xvars1)),
                    effective_df=rep(effective_df,each=length(Xvars1)))
nvars <- length(Xvars1)

for (i in 1:length(Lambda)) {
  m_tmp <- lmm_fit("ln_art_utilisation1",Xvars1,"country",art_dat[exclude==FALSE],lambda=Lambda[i])
  betas$beta[((i-1)*nvars + 1):(i*nvars)] <- m_tmp$beta
}
setDT(betas)
```

The relationship between the penalty and effective degrees of freedom.

```{r}
ggplot(betas,aes(x=lambda,y=effective_df)) +
  geom_point() +
  theme_bw(base_size=12) +
  labs(x=expression(lambda),y="Effective degrees of freedom")
```

```{r, fig.height=4, fig.width=8}
sec_el <- seq(1, length(Lambda), 2)
ggplot(betas,aes(x=(lambda),y=beta,col=var)) + 
  geom_line(size=1) +
  geom_hline(yintercept = 0.0) +
  geom_vline(xintercept = 100.0,linetype=2) +
  theme_bw(base_size=12) +
  scale_x_continuous(breaks=Lambda[sec_el],labels = round(effective_df,2)[sec_el]) + 
  scale_color_discrete(labels=Xvars1_labels,name="") +
  theme(axis.text.x = element_text(angle=45,vjust=0.5),panel.grid=element_blank()) +
  labs(x="Effective number of parameters",y=expression(beta))
```

Then we refit the penalised model. 

**These results are for the component sensitivity analysis (full cases) model**

```{r}
mc01p <- lmm_fit("ln_art_utilisation1",Xvars1,"country",art_dat[exclude==FALSE],lambda=75.0)
saveRDS(mc01p,file="../models/component_standard_l275_complete.rds")
mc01p_confint <- data.frame(beta=mc01p$beta,
                            se=mc01p$beta_se,
                            lower=mc01p$beta-2*mc01p$beta_se,
                            upper=mc01p$beta+2*mc01p$beta_se)
knitr::kable(mc01p_confint,digits=3,caption = "Model 95% intervals")
fwrite(x = as.data.table(mc01p_confint,keep.rownames = TRUE),file = "../results/tables/component_standard_l275_complete.csv")
```

The marginal R^2

```{r}
marginal_r2(mc01p,art_dat,Xvars1,art_dat$ln_art_utilisation1)
```

## Model for each component

### Adol Birth Rate

```{r}
# model
model <- lmer(ln_art_utilisation1 ~ 1+
               time1+
               ln_adol_birth_rate_wb1+
               (1|country),
             data = art_dat[exclude==FALSE],
             REML=FALSE)
cat("R2: ",MuMIn::r.squaredGLMM(model),"\n")
saveRDS(model,file="../models/component_ln_adol_birth_rate_wb1_complete.rds")

# model details
augment_model <- augment(model)
setDT(augment_model)

# model fit
knitr::kable(tidy(model),digits = 3,caption = "Model parameter estimates")
knitr::kable(confint(model),digits=3,caption = "Model 95% intervals")
knitr::kable(glance(model),digits = 3,caption = "Model fit statistics")
```

### ln_maternal_mortality_wb

```{r}
# model
model <- lmer(ln_art_utilisation1 ~ 1+
               time1+
               ln_maternal_mortality_wb1+
               (1|country),
             data = art_dat[exclude==FALSE],
             REML=FALSE)
cat("R2: ",MuMIn::r.squaredGLMM(model),"\n")
saveRDS(model,file="../models/component_ln_maternal_mortality_wb1_complete.rds")

# model details
augment_model <- augment(model)
setDT(augment_model)

# model fit
knitr::kable(tidy(model),digits = 3,caption = "Model parameter estimates")
knitr::kable(confint(model),digits=3,caption = "Model 95% intervals")
knitr::kable(glance(model),digits = 3,caption = "Model fit statistics")
```

### women_parl_wb

```{r}
# model
model <- lmer(ln_art_utilisation1 ~ 1+
               time1+
               women_parl_wb1+
               (1|country),
             data = art_dat[exclude==FALSE],
             REML=FALSE)
cat("R2: ",MuMIn::r.squaredGLMM(model),"\n")
saveRDS(model,file="../models/component_women_parl_wb1_complete.rds")

# model details
augment_model <- augment(model)
setDT(augment_model)

# model fit
knitr::kable(tidy(model),digits = 3,caption = "Model parameter estimates")
knitr::kable(confint(model),digits=3,caption = "Model 95% intervals")
knitr::kable(glance(model),digits = 3,caption = "Model fit statistics")
```

### school_gap

```{r}
# model
model <- lmer(ln_art_utilisation1 ~ 1+
               time1+
               school_gap1+
               (1|country),
             data = art_dat[exclude==FALSE],
             REML=FALSE)
cat("R2: ",MuMIn::r.squaredGLMM(model),"\n")
saveRDS(model,file="../models/component_school_gap1_complete.rds")

# model details
augment_model <- augment(model)
setDT(augment_model)

# model fit
knitr::kable(tidy(model),digits = 3,caption = "Model parameter estimates")
knitr::kable(confint(model),digits=3,caption = "Model 95% intervals")
knitr::kable(glance(model),digits = 3,caption = "Model fit statistics")
```

### labour_gap_wb

```{r}
# model
model <- lmer(ln_art_utilisation1 ~ 1+
               time1+
               labour_gap_wb1+
               (1|country),
             data = art_dat[exclude==FALSE],
             REML=FALSE)
cat("R2: ",MuMIn::r.squaredGLMM(model),"\n")
saveRDS(model,file="../models/component_labour_gap_wb1_complete.rds")

# model details
augment_model <- augment(model)
setDT(augment_model)

# model fit
knitr::kable(tidy(model),digits = 3,caption = "Model parameter estimates")
knitr::kable(confint(model),digits=3,caption = "Model 95% intervals")
knitr::kable(glance(model),digits = 3,caption = "Model fit statistics")
```

### mean_sch_un

```{r}
# model
model <- lmer(ln_art_utilisation1 ~ 1+
               time1+
               mean_sch_un1+
               (1|country),
             data = art_dat[exclude==FALSE],
             REML=FALSE)
cat("R2: ",MuMIn::r.squaredGLMM(model),"\n")
saveRDS(model,file="../models/component_mean_sch_un1_complete.rds")

# model details
augment_model <- augment(model)
setDT(augment_model)

# model fit
knitr::kable(tidy(model),digits = 3,caption = "Model parameter estimates")
knitr::kable(confint(model),digits=3,caption = "Model 95% intervals")
knitr::kable(glance(model),digits = 3,caption = "Model fit statistics")
```

### life_expect_wb

```{r}
# model
model <- lmer(ln_art_utilisation1 ~ 1+
               time1+
               life_expect_wb1+
               (1|country),
             data = art_dat[exclude==FALSE],
             REML=FALSE)
cat("R2: ",MuMIn::r.squaredGLMM(model),"\n")
saveRDS(model,file="../models/component_life_expect_wb1_complete.rds")

# model details
augment_model <- augment(model)
setDT(augment_model)

# model fit
knitr::kable(tidy(model),digits = 3,caption = "Model parameter estimates")
knitr::kable(confint(model),digits=3,caption = "Model 95% intervals")
knitr::kable(glance(model),digits = 3,caption = "Model fit statistics")
```

### ln_gni_wb

```{r}
# model
model <- lmer(ln_art_utilisation1 ~ 1+
               time1+
               ln_gni_wb1+
               (1|country),
             data = art_dat[exclude==FALSE],
             REML=FALSE)
cat("R2: ",MuMIn::r.squaredGLMM(model),"\n")
saveRDS(model,file="../models/component_ln_gni_wb1_complete.rds")

# model details
augment_model <- augment(model)
setDT(augment_model)

# model fit
knitr::kable(tidy(model),digits = 3,caption = "Model parameter estimates")
knitr::kable(confint(model),digits=3,caption = "Model 95% intervals")
knitr::kable(glance(model),digits = 3,caption = "Model fit statistics")
```

# Appendix A: additional models

## Standard formulation with excluded countries

A better model with the excluded countries included adds a Middle East/North Africa
indicator variable. Note that due to the high variability of (within country) ART rates in the excluded
countries these results may not be reliable. 

Model: log ART utilisation = time in years + GII + HDI + Middle East/North Africa + random country intercept

```{r}
# model
mm01a <- lmer(ln_art_utilisation ~ 1+time+gii_un+hdi_un+me_na+(1|country), data = art_dat)
cat("R2: ",MuMIn::r.squaredGLMM(mm01a),"\n")

# model details
augment_mm01a <- augment(mm01a)
setDT(augment_mm01a)

# model fit
knitr::kable(tidy(mm01a),digits = 3,caption = "Model parameter estimates")
knitr::kable(glance(mm01a),digits = 3,caption = "Model fit statistics")
```

## Decomposition formulation with excluded countries

Note that due to the high variability of (within country) ART rates in the excluded
countries these results may not be reliable. 

```{r}
# model
mm02a <- lmer(ln_art_utilisation ~ 1+time+gii_d+gii_m+hdi_d+hdi_m+me_na+(1|country),data = art_dat)
cat("R2: ",MuMIn::r.squaredGLMM(mm02a),"\n")

# model details
augment_mm02a <- augment(mm02a)
setDT(augment_mm02a)

# model fit
knitr::kable(tidy(mm02a),digits = 3,caption = "Model parameter estimates")
knitr::kable(glance(mm02a),digits = 3,caption = "Model fit statistics")
```

## Standard formulation with Turkey excluded

Turkey was an influential outlier above - effect of removing.

```{r}
# model
mm01a <- lmer(ln_art_utilisation ~ 1+time+gii_un+hdi_un+me_na+(1|country), 
              data = art_dat[!country=="Turkey" & exclude == FALSE])
cat("R2: ",MuMIn::r.squaredGLMM(mm01a),"\n")

# model details
augment_mm01a <- augment(mm01a)
setDT(augment_mm01a)

# model fit
knitr::kable(tidy(mm01a),digits = 3,caption = "Model parameter estimates")
knitr::kable(glance(mm01a),digits = 3,caption = "Model fit statistics")
```

## Decomposition formulation with Turkey excluded

Turkey was an influential outlier above - effect of removing.

```{r}
# model
mm02a <- lmer(ln_art_utilisation ~ 1+time+gii_d+gii_m+hdi_d+hdi_m+me_na+(1|country),
              data = art_dat[!country=="Turkey" & exclude == FALSE])
cat("R2: ",MuMIn::r.squaredGLMM(mm02a),"\n")

# model details
augment_mm02a <- augment(mm02a)
setDT(augment_mm02a)

# model fit
knitr::kable(tidy(mm02a),digits = 3,caption = "Model parameter estimates")
knitr::kable(glance(mm02a),digits = 3,caption = "Model fit statistics")
```


