---
title: "Model for ART utilisation over time"
subtitle: "Complete case analysis"
author: "Oisin Fitzgerald"
output:
  html_document:
    toc: true
---

```{r,echo=FALSE}
knitr::opts_chunk$set(echo=FALSE,message=FALSE,warning=FALSE)
```

# Note

The file `scripts/impute.R` needs to be be run before this file. However,
knitting this file before running the imputations will result in `scripts/impute.R`
being sourced (but you'll have a long wait). 

# Introduction

This document describes the longitudinal modeling of assisted reproductive 
technology utilisation (ART) as a function of the United Nations produced 
Human Development Index (HDI) and Gender Inequality Index (GII). Greater details on the 
rationale for the study can be found in the main manuscript. 

Summary of methods and findings:

* Linear mixed effect model for log ART over time with covariates GII and HDI  
* GII only available for year XXX so data restricted to these years.
* GII related to log ART, increased GII related to increased ART utilisation. Effect marginally significant.

```{r}
impute_files <- paste0("../data/imputations/",
                       c("implist11.RData","implist22.RData","implist33.RData"))
if (!all(file.exists(impute_files))) {
  source("../scripts/impute/impute.R")
}
```

```{r doc_setup}
### packages, functions
# external
library(lme4)
library(mitml)
library(MuMIn)
library(ggplot2)
library(data.table)

# internal
source("../src/tools/visualise.R")
source("../src/tools/summarise.R")
source("../src/tools/modelling.R")
source("../src/tools/gii_comp.R")
source("../src/lmm_ridge.R")

### import data
load("../data/imputations/implist11.RData")  # partial missing data imputated
load("../data/imputations/implist22.RData")  # all missing data imputated
load("../data/imputations/implist33.RData")  # pre 2009 missing data imputated
```

```{r}
## function for preparing the imputed datasets
scale <- function(x) (x - mean(x,na.rm=TRUE))/sd(x,na.rm = TRUE)
preprocess_data <- function(x) {
  # some renaming
  names(x)[names(x) == "ln_adol_birth_rate_wb"] <- "ln_abr"
  names(x)[names(x) == "ln_maternal_mortality_wb"] <- "ln_mmr"

  ## missing reponse
  incl <- !is.na(x$ln_art_utilisation)
  # time
  x$time_m <- as.numeric(tapply(x$time[incl],x$country[incl], mean))[x$country]
  x$time_d <- x$time - x$time_m
  # indices
  x$gii <- calc_gii(exp(x$ln_mmr),
                      exp(x$ln_abr),
                      100*(x$female_lfp_wb),
                      100*(x$male_lfp_wb),
                      100*(x$EA_2T8_AG25T99_F),
                      100*(x$EA_2T8_AG25T99_M),
                      100*(x$women_parl_wb))
  x$gii_m <- as.numeric(tapply(x$gii[incl],x$country[incl], mean))[x$country]
  x$gii_d <- x$gii - x$gii_m
  x$hdi_m <- as.numeric(tapply(x$hdi_un[incl],x$country[incl], mean))[x$country]
  x$hdi_d <- x$hdi_un - x$hdi_m
  # create labour and school gap
  x$sch_gap <- x$EA_2T8_AG25T99_M - x$EA_2T8_AG25T99_F
  # x$lab_gap_ilo <- x$lfp_male_ilo - x$lfp_female_ilo
  x$lab_gap_wb <- x$male_lfp_wb - x$female_lfp_wb
  # changes in components
  x$sch_gap_m <- as.numeric(tapply(x$sch_gap[incl],x$country[incl], mean))[x$country]
  x$sch_gap_d <- x$sch_gap - x$sch_gap_m
  x$sch_gap_d2 <- x$sch_gap_d
  x$sch_gap_d2[x$sch_gap_d2 < 0] <- 0
  # x$lab_gap_ilo_m <- as.numeric(tapply(x$lab_gap_ilo[incl],x$country[incl], mean))[x$country]
  # x$lab_gap_ilo_d <- x$lab_gap_ilo - x$lab_gap_ilo_m
  # x$lab_gap_ilo_d2 <- x$lab_gap_ilo_d
  # x$lab_gap_ilo_d2[x$lab_gap_ilo_d2 < 0] <- 0
  x$lab_gap_wb_m <- as.numeric(tapply(x$lab_gap_wb[incl],x$country[incl], mean))[x$country]
  x$lab_gap_wb_d <- x$lab_gap_wb - x$lab_gap_wb_m
  x$women_parl_wb_m <- as.numeric(tapply(x$women_parl_wb[incl],x$country[incl], mean))[x$country]
  x$women_parl_wb_d <- x$women_parl_wb - x$women_parl_wb_m
  x$ln_abr_m <- as.numeric(tapply(x$ln_abr[incl],x$country[incl], mean))[x$country]
  x$ln_abr_d <- x$ln_abr - x$ln_abr_m
  x$ln_mmr_m <- as.numeric(tapply(x$ln_mmr[incl],x$country[incl], mean))[x$country]
  x$ln_mmr_d <- x$ln_mmr - x$ln_mmr_m
  x$ln_gni_m <- as.numeric(tapply(x$ln_gni_wb[incl],x$country[incl], mean))[x$country]
  x$ln_gni_d <- x$ln_gni_wb - x$ln_gni_m
  # health-wealth-edu
  x$hwe <- 0.3*x$ln_gni_wb+0.8*x$mean_sch_un+0.3*x$ln_abr-0.4*x$ln_mmr
  x$hwe_m <- as.numeric(tapply(x$hwe[incl],x$country[incl], mean))[x$country]
  x$hwe_d <- x$hwe - x$hwe_m
  # outlier
  x$out1 <- rep(0, nrow(x))
  x$out1[x$sch_gap > 0.2] <- 1
  # scale some data
  x$time1  <- scale(x$time)
  x$ln_abr1 <- scale(x$ln_abr)
  x$women_parl_wb1 <- scale(x$women_parl_wb)
  x$ln_mmr1 <- scale(x$ln_mmr)
  x$sch_gap1 <- scale(x$sch_gap)
  x$lab_gap_wb1 <- scale(x$lab_gap_wb)
  x$life_expect_wb1 <- scale(x$life_expect_wb)
  x$ln_gni_wb1 <- scale(x$ln_gni_wb)
  x$mean_sch_un1 <- scale(x$mean_sch_un)
  x$gii <- scale(x$gii)
  x$hdi_un <- scale(x$hdi_un)
  x$time <- scale(x$time)
  x$ln_art_utilisation1 <- x$ln_art_utilisation - mean(x$ln_art_utilisation,na.rm=TRUE)
  # return data
  x
}

implist11 <- lapply(implist11, preprocess_data)
class(implist11) <- c("mitml.list","list")
implist22 <- lapply(implist22, preprocess_data)
class(implist22) <- c("mitml.list","list")
implist33 <- lapply(implist33, preprocess_data)
class(implist33) <- c("mitml.list","list")
```

# Index Models

## Covariate imputation

Covariates imputed.

### Standard formulation

```{r}
# standard random effects
mmpi01 <- with(implist11, {
  lmer(ln_art_utilisation~1+time+gii+hdi_un+(1|country),REML=TRUE)
})
res_mmpi01 <- testEstimates(mmpi01, extra.pars=TRUE)
saveRDS(res_mmpi01,file="../models/index_standard_partial.rds")

# parameter estimates
knitr::kable(res_mmpi01$estimates,digits=3,
             caption="Model parameter estimates")

# confidence intervals
knitr::kable(confint(res_mmpi01),digits=3,caption="Model parameter 95% CI")

# variance components
knitr::kable(res_mmpi01$extra.pars,digits=3,
             caption="Variance component estimates")

tab <- cbind(res_mmpi01$estimates,confint(res_mmpi01))
fwrite(x=as.data.table(tab,keep.rownames = TRUE),file="../results/tables/index_standard_partial.csv")
```

### Decomposition formulation

```{r}
# between-within decomposition
mmpi02 <- with(implist11, {
  lmer(ln_art_utilisation~1+
         time_d+
         time_m+
         gii_d+
         gii_m+
         hdi_d+
         hdi_m+
         (1|country),REML=TRUE)
})
res_mmpi02 <- testEstimates(mmpi02, extra.pars=TRUE)
saveRDS(res_mmpi02,file="../models/index_decomposition_partial.rds")

# parameter estimates
knitr::kable(res_mmpi02$estimates,digits=3,
             caption="Model parameter estimates")

# confidence intervals
knitr::kable(confint(res_mmpi02),digits=3,caption="Model parameter 95% CI")

# variance components
knitr::kable(res_mmpi02$extra.pars,digits=3,
             caption="Variance component estimates")
```

GII only

```{r}
# between-within decomposition
model <- with(implist11, {
  lmer(ln_art_utilisation~1+
         time+
         gii_d+
         gii_m+
         (1|country),REML=TRUE)
})
res_model <- testEstimates(model, extra.pars=TRUE)

# parameter estimates
knitr::kable(res_model$estimates,digits=3,
             caption="Model parameter estimates")

# confidence intervals
knitr::kable(confint(res_model),digits=3,caption="Model parameter 95% CI")

# variance components
knitr::kable(res_model$extra.pars,digits=3,
             caption="Variance component estimates")
```

## Covariates and outcome imputation

### Standard formulation

```{r}
# standard random effects
mmfi01 <- with(implist22, {
  lmer(ln_art_utilisation~1+
         time+
         gii+
         hdi_un+
         (1|country),REML=TRUE)
})
res_mmfi01 <- testEstimates(mmfi01, extra.pars=TRUE)
saveRDS(res_mmfi01,file="../models/index_standard_full.rds")

# parameter estimates
knitr::kable(res_mmfi01$estimates,digits=3,
             caption="Model parameter estimates")

# confidence intervals
knitr::kable(confint(res_mmfi01),digits=3,caption="Model parameter 95% CI")

# variance components
knitr::kable(res_mmfi01$extra.pars,digits=3,
             caption="Variance component estimates")
             
tab <- cbind(res_mmfi01$estimates,confint(res_mmfi01))
fwrite(x=as.data.table(tab,keep.rownames = TRUE),file="../results/tables/index_standard_full.csv")

```

### Decomposition formulation

GII + HDI

```{r}
# between-within decomposition
mmfi02 <- with(implist22, {
  lmer(ln_art_utilisation~1+
         time+
         gii_d+
         gii_m+
         hdi_d+
         hdi_m+
         (1|country),REML=TRUE)
})
res_mmfi02 <- testEstimates(mmfi02, extra.pars=TRUE)
saveRDS(res_mmfi02,file="../models/index_decomposition_full.rds")

# parameter estimates
knitr::kable(res_mmfi02$estimates,digits=3,
             caption="Model parameter estimates")

# confidence intervals
knitr::kable(confint(res_mmfi02),digits=3,caption="Model parameter 95% CI")

# variance components
knitr::kable(res_mmfi02$extra.pars,digits=3,
             caption="Variance component estimates")
```

GII only

```{r}
# between-within decomposition
model <- with(implist22, {
  lmer(ln_art_utilisation~1+
         time+
         gii_d+
         gii_m+
         (1|country),REML=TRUE)
})
res_model <- testEstimates(model, extra.pars=TRUE)

# parameter estimates
knitr::kable(res_model$estimates,digits=3,
             caption="Model parameter estimates")

# confidence intervals
knitr::kable(confint(res_model),digits=3,caption="Model parameter 95% CI")

# variance components
knitr::kable(res_model$extra.pars,digits=3,
             caption="Variance component estimates")
```

# Component Models

## Covariate imputation

### Standard formulation - no shrinkage on fixed effects

#### Unscaled

```{r}
# model
mmcpi01 <- with(implist11,
               lmer(ln_art_utilisation ~ 
                 1+time+
               ln_abr+
               ln_mmr+
               women_parl_wb+
               sch_gap+
                 mean_sch_un+
               lab_gap_wb+
                life_expect_wb+
               ln_gni_wb+
               (1|country)))
res_mmcpi01 <- testEstimates(mmcpi01, extra.pars=TRUE)

# parameter estimates
knitr::kable(res_mmcpi01$estimates,digits=3,
             caption="Model parameter estimates")

# confidence intervals
knitr::kable(confint(res_mmcpi01),digits=3,caption="Model parameter 95% CI")

# variance components
knitr::kable(res_mmcpi01$extra.pars,digits=3,
             caption="Variance component estimates")
```

### Scaled

```{r}
# model
mmcpi01s <- with(implist11,
               lmer(ln_art_utilisation1 ~ -1+
               time1+
               ln_abr1+
               ln_mmr1+
               women_parl_wb1+
               sch_gap1+
               lab_gap_wb1+
               mean_sch_un1+
               life_expect_wb1+
               ln_gni_wb1+
               (1|country)))
res_mmcpi01s <- testEstimates(mmcpi01s, extra.pars=TRUE)

# parameter estimates
knitr::kable(res_mmcpi01s$estimates,digits=3,
             caption="Model parameter estimates")

# confidence intervals
knitr::kable(confint(res_mmcpi01s),digits=3,caption="Model parameter 95% CI")

# variance components
knitr::kable(res_mmcpi01s$extra.pars,digits=3,
             caption="Variance component estimates")
```

### Standard formulation - L2 shrinkage on fixed effects

```{r}
Xvars1 <- c("time1","ln_abr1", "women_parl_wb1", "ln_mmr1",
          "sch_gap1", "lab_gap_wb1", "life_expect_wb1", "ln_gni_wb1", "mean_sch_un1")
Xvars1_labels <- c("Year","Adol. birth rate","Women in parliament (%)","Maternal mort. ratio",
                   "Gender highschool gap","Gender LFP gap","Life expectancy",
                   "log GNI","Av. years of schooling")
art_dat <- implist11[[1]]
Lambda <- seq(1,1500,length.out = 20)
effective_df <- numeric(length=length(Lambda))
X <- as.matrix(na.omit(art_dat[,..Xvars1]))
for (i in 1:length(effective_df)) {
  effective_df[i] <- sum(diag(X %*% solve(t(X) %*% X + diag(Lambda[i],length(Xvars1))) %*% t(X)))
}

betas <- data.frame(beta=numeric(length = length(Lambda)*length(Xvars1)),
                    var=rep(Xvars1,length(Lambda)),
                    lambda=rep(Lambda,each=length(Xvars1)),
                    effective_df=rep(effective_df,each=length(Xvars1)))
nvars <- length(Xvars1)

for (i in 1:length(Lambda)) {
  m_tmp <- lmm_fit("ln_art_utilisation1",Xvars1,"country",art_dat,lambda=Lambda[i])
  betas$beta[((i-1)*nvars + 1):(i*nvars)] <- m_tmp$beta
}
setDT(betas)
```

```{r}
ggplot(betas,aes(x=lambda,y=effective_df)) +
  geom_point() +
  theme_bw(base_size=12) +
  labs(x=expression(lambda),y="Effective degrees of freedom")
```

```{r, fig.height=4, fig.width=8}
sec_el <- seq(1, length(Lambda), 2)
ggplot(betas,aes(x=(lambda),y=beta,col=var)) + 
  geom_line(size=1) +
  geom_hline(yintercept = 0.0) +
  geom_vline(xintercept = 75.0,linetype=2) +
  theme_bw(base_size=12) +
  scale_x_continuous(breaks=Lambda[sec_el],labels = round(effective_df,2)[sec_el]) + 
  scale_color_discrete(labels=Xvars1_labels,name="") +
  theme(axis.text.x = element_text(angle=45,vjust=0.5),panel.grid=element_blank()) +
  labs(x="Effective number of parameters",y=expression(beta))
```

Only consider scaled, else interpretation confusing.

**These results are for the component analysis main model**

```{r}
# model
f <- function(x) {
  mod <- lmm_fit("ln_art_utilisation1",Xvars1,"country",x,lambda=75.0)
  mod
}
mmcpi01s <- lapply(implist11,f)

beta <- lapply(mmcpi01s,function(x) x$beta)
beta_var <- lapply(mmcpi01s,function(x) x$beta_se^2)
res_mmcpi01s <- testEstimates(qhat=beta,uhat=beta_var)
saveRDS(res_mmcpi01s,file="../models/components_standard_l275_partial.rds")

# parameter estimates
knitr::kable(res_mmcpi01s$estimates,digits=3,
             caption="Model parameter estimates")

# confidence intervals
knitr::kable(confint(res_mmcpi01s),digits=3,caption="Model parameter 95% CI")
```

```{r}
tab <- cbind(res_mmcpi01s$estimates,confint(res_mmcpi01s))
fwrite(x=as.data.table(tab,keep.rownames = TRUE),file="../results/tables/component_standard_l275_partial.csv")
```


## Covariates and outcome imputation

### Standard formulation - L2 shrinkage on fixed effects

**These results are for the component analysis (sensitivity analysis, full imputation)**

```{r}
# model
f <- function(x) {
  mod <- lmm_fit("ln_art_utilisation1",Xvars1,"country",x,lambda=75.0)
  mod
}
mmcfi01s <- lapply(implist22,f)

beta <- lapply(mmcfi01s,function(x) x$beta)
beta_var <- lapply(mmcfi01s,function(x) x$beta_se^2)
res_mmcfi01s <- testEstimates(qhat=beta,uhat=beta_var)
saveRDS(res_mmcfi01s,file="../models/components_standard_l275_full.rds")

# parameter estimates
knitr::kable(res_mmcfi01s$estimates,digits=3,
             caption="Model parameter estimates")

# confidence intervals
knitr::kable(confint(res_mmcfi01s),digits=3,caption="Model parameter 95% CI")

```


```{r}
tab <- cbind(res_mmcfi01s$estimates,confint(res_mmcfi01s))
fwrite(x=as.data.table(tab,keep.rownames = TRUE),file="../results/tables/component_standard_l275_full.csv")
```



